WSClean: a new implementation of a fast, generic wide-field imager

### 2014-01-23 articles/2014/Article-WSClean/wsclean.tex
Todays work on article

To derive the w-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \cdot \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \cdot \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \cdot \\ \notag
& e^{-2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_0$ to $w_e$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_e - w_0\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \cdot \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{-2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~taking all samples that correspond to a specific value of $w$; ii)~grid those samples on a uniform grid; iii)~calculating the inverse FFT; iv)~apply the direction dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; v)~repeat this for all $w$-values and add the results together; vi)~apply the final scaling.

@@

@@

\subsection{Full-sky imaging}
For low-frequency telescopes, it can be very useful to do full-sky imaging. In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90 degrees from the pointing centre. Imaging these sidelobes can be useful because there might be science (e.g. when searching for transients), or for self-calibration or full-sky deconvolution. The latter is for example useful when a strong source in a sidelobe affects the sensitivity in the pointing centre.

\subsection{W-snapshots}
TODO discuss projecting visibilities onto different tangent plane, imaging by small change to algorithm, output standard fits file by using CPIXij keyword \citet{wcs-in-fits}.
, and the w-terms can be kept small by phasing a snapshot to the orthogonal of the array's best-fitting plane

\section{The WSClean imager implementation} \label{sec:implementation}
We have implemented the $w$-stacking algorithm in the C++ language. Several steps are multi-threaded in this implementation: reading and writing, gridding from and to different $w$-layers, performing the Fast Fourier transforms and peak-finding and image subtraction during Cleaning. For the latter, intrinsics were used as well. This speeds up a minor Cleaning iterations,

@@

>>>>

To derive the w-stacking technique, Eq.~\eqref{eq:visibility-function} is rewritten to
\begin{align}\notag
V(u,v,w) = & \iint \frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} \cdot \\ \notag
& e^{-2\pi i \left(ul + vm\right)} dl dm.
\end{align}
This is an ordinary two-dimensional Fourier transform going from $u,v$ space to $l,m$ space, and can be inverted to get:
% Intermediate step:
%\begin{align}\notag
%\frac{I'(l,m) e^{-2\pi i w(\sqrt{1-l^2-m^2}-1)}}{\sqrt{1-l^2-m^2}} = & \iint V(u,v,w) \cdot \\ \notag
%& e^{-2\pi i \left(ul + vm\right)} du dv,
%\end{align}
%which can be rewritten to
\begin{align}\notag
\frac{I'(l,m)}{\sqrt{1-l^2-m^2}} = & e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \iint V(u,v,w) \cdot \\ \notag
& e^{2\pi i \left(ul + vm\right)} du dv.
\end{align}
Integrating both sides over $w_0$ to $w_e$, the minimum and maximum value of $w$, results in
\begin{align}\notag
\frac{I'(l,m)\left(w_e - w_0\right)}{\sqrt{1-l^2-m^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-l^2-m^2}-1)} \cdot \\ \label{eq:wstacking}
\iint V(u,v,w)  e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
The final step is to make the $u,v,w$ parameters discrete, so that the integration over $u$ and $v$ can become an inverse FFT and the integration over $w$ becomes a summation. This shows that the sky function can be reconstructed by i)~taking all samples that correspond to a specific value of $w$; ii)~grid those samples on a uniform grid; iii)~calculating the inverse FFT; iv)~apply the direction dependent phaseshift $e^{2\pi i w(\sqrt{1-l^2-m^2}-1)}$; v)~repeat this for all $w$-values and add the results together; vi)~apply the final scaling.

@@

TODO: discuss visibility prediction? (inverse of imaging)

@@

\section{The WSClean imager implementation} \label{sec:implementation}
We have implemented the $w$-stacking algorithm in the C++ language. Several steps are multi-threaded in this implementation: reading and writing; gridding from and to different $w$-layers; performing the Fast Fourier transforms; and minor Clean iterations (peak-finding and image subtraction). For the latter, intrinsics were used as well. Because of these optimizations, minor Cleaning iterations are so fast that performing the Clark Clean optimization \citep{clark-clean} of considering only a subset of pixels with a reduced PSF, is not so relevant. However, because the PSF varies for imaging with a non-zero w-value, major iterations are still required to fix the errors made by subtracting a constant PSF during minor iterations.

@@

\subsection{Full-sky imaging}
For low-frequency telescopes, it can be useful to do full-sky imaging. In the case of MWA, the tile beam can have strong sidelobes at a distance of more than 90 degrees from the pointing centre. Imaging these sidelobes might be relevant because it might either be scientifically interesting (e.g. when searching for transients), or be useful for self-calibration or full-sky deconvolution. For example, the latter is useful when a strong source in a sidelobe affects the sensitivity near the pointing centre.

To image the full sky, an observation is split into short snapshots, their phase centre is changed to zenith and subsequently a suitable area is imaged. Depending on desired resolution and dynamic range, this might require very large images. To make an image at the MWA resolution, the image needs to have approximately 10K pixels in width and height. The $w$-values will be very small at zenith, because only the differential heights of the antennas will contribute to $w$-value at zenith. For the MWA, the maximum height difference between tiles is $9.3$m. The MWA tiles are in fact on a slight slope, and by fitting a plane to the antennas and changing the phase centre towards the normal of that plane, the $w$-value decreases to $7.2\textrm{m} / \lambda$.

(TODO: these $w$ values are in 2 min obses; need shorter time interval for correct values).

\subsection{W-snapshots}
To get optimal imaging performance, the $w$-values can be kept small by phase-changing towards zenith. As shown, all-sky imaging can be quite efficient with this technique. However, if one is not interested in imaging the whole sky, and the direction of interest is far from zenith, the computational overhead of making all-sky images is undesirable.

Instead of imaging the whole sky, the image can be recentred from $(l,m)$ to $(\hat{l},\hat{m})=(l+\Delta l,m+\Delta m)$ by performing the substitution $(l,m)\leftarrow(\hat{l},\hat{m})$ in Eq.~\eqref{eq:wstacking}:
%When the Fourier transform is performed, the $w$ term has been corrected (either by w-stacking or by w-projection). Therefore, if $U(u, v)$ is the input to the 2-dimensional Fourier transform, the transform is described by
\begin{align}\notag
\frac{I'(\hat{l}, \hat{m})\left(w_e - w_0\right)}{\sqrt{1-\hat{l}^2-\hat{m}^2}} = \int\limits_{w_0}^{w_e} e^{2\pi i w(\sqrt{1-\hat{l}^2-\hat{m}^2}-1)} \cdot \\
\iint e^{2 \pi i \left(u\Delta l + v\Delta m\right)} V(u,v,w) e^{2\pi i \left(ul + vm\right)} du dv dw.
\end{align}
In words, recentering an image involves accounting for the position shift during the $w$-correction and final scaling, and shifting the visibilities in phase by multiplication with $e^{2 \pi i \left(u\Delta l + v\Delta m\right)}$ prior to the imaging.

A recentred image is in the projection of the zenith tangent plane, and will need an additional regridding step before multiple snapshots can be added together.

Another method to shift an image, is by using the periodicity of the FFT function. Since sources outside the field of view will be aliased back into the field, the image can be regridded with the alias in the centre. This complicates gridding, because the antialiasing kernel needs to have a pass-band shape instead of a low-pass shape. Therefore, we chose to implement the former method.

TODO discuss output standard fits file by using CPIXij keyword \citet{wcs-in-fits}.
